\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}

### 2022-05-10 lin-vs-cen.tex
Cleaned up some calculations

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4)$ such that the following is true, for every integer $k\ge 1$.\newline
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

@@

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{r}$ for each $\alpha\in X$.
  For each $\alpha\in X$, let $\Phi_\alpha:=\{\nu\in\VE(G_k):\phi(\nu)=\alpha\}$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\Phi_\alpha| \ge d$.  Furthermore all but at most $\vol{r}$ elements of $\Phi_\alpha$ are contained in $\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))$ since otherwise the first round would have been an $\alpha$-success.\todo{Explain more or redefine $\Phi_\alpha$ as $\Phi_\alpha\cap (\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu)))$)}
  % For each $\mu\in Y$, $|\VE(G_{r+1}(\mu))|\in O(r^2)$. Therefore, there are $\Omega(|\Phi_\alpha|/r^2)$ objects $\mu\in Y$ such that $\VE(G_{r+1}(\mu))\cap \Phi_\alpha\neq\emptyset$.

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\ceil{|\Phi_\alpha|/\vol{10r}},1/25)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{r})$.  (We make no attempt to optimize constants here.)  Suppose that, for some $\mu\in Y$, $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{2r}(\xi))\}
  \]
  has size at most $25$. Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/25$. This already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$.

  In order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_r(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J-\VE(G_{10r}(\mu))$.\footnote{The constant $10$ is overkill here and is only used for simplicity; $8r+O(1)$ would be sufficient.}  This process continues until $\VE(G_r(\mu))\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{10r}(\mu))|\le \vol{10r}$, each step in this process eliminates at most $\vol{10r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{10r}}$

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/25)\le 2\vol{r}) \notag \\
    & = \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{1}{25}\right)^x\left(\frac{24}{25}\right)^{t-x} \notag \\
    & < \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{24}{25}\right)^{t} \notag \\
    & \le (2t)^{2\vol{r}}\cdot\left(\frac{24}{25}\right)^{t} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & = \exp\big({2\vol{r}}\ln(2t)-t\ln(25/24)\big) \notag \\
    & \le \exp\left({2\vol{r}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|\ln(25/24)}{\vol{10r}}\right)
      & \text{(since $|\Phi_\alpha|/\vol{10r} \le t \le |\Phi_\alpha|$)} \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

  We are now ready for an application of \cref{weighted_lovasz}.  For each $\alpha\in X$, let Let $E_\alpha$ denote the event ``$\deg_H(\alpha)\le 2\vol{r}$'' and let $\mathcal{E}:=\{E_\alpha:\alpha\in X\}$. For each $\alpha\in X$, define
  \[
     \Gamma_\alpha := \left\{E_\beta \in \mathcal{E}: \Phi_\beta\cap \bigcup_{\nu\in \Phi_\alpha} \VE(G_{10r}(\nu))\neq\emptyset\right\}
  \]
  and observe that $|\Gamma_\alpha|\le \vol{10r}\cdot|\Phi_\alpha|$.
  The event $E_\alpha$ is mutually independent of $\mathcal{E}\setminus\Gamma_\alpha$ since, for any $E_\beta\in\mathcal{E}\setminus\Gamma_\alpha$ the sets $Y_\alpha:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\alpha\neq\emptyset\}$ and $Y_\beta:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\beta\neq\emptyset\}$ are disjoint and $\deg_H(\alpha)$ and $\deg_H(\beta)$ are each determined by the relative orders of $Y_\alpha$ and $Y_\beta$ in $\pi$, respectively.

  For each $\alpha\in X$, let $w(E_\alpha):=1/\tau$, for some $\tau>0$ to be determined shortly.  Then
  \begin{align}
    w(E_\alpha)\prod_{E_\beta\in\Gamma_\alpha} (1-w(E_\beta))
    & = \tfrac{1}{\tau}\left(1-\tfrac{1}{\tau}\right)^{|\Gamma_\alpha|-1} \notag \\
    & \ge \exp\left(-\log\tau - |\Gamma_\alpha|/(\tau-1)\right) & \text{(Since $(1-1/\tau)^{\tau-1} > 1/e$)} \notag \\
    & \ge \exp\left(-\log\tau - \vol{10r}\cdot|\Phi_\alpha|/(\tau-1)\right) \enspace . \label{weight_product}
  \end{align}

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {50\vol{10r}}+1$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(\Gamma_\alpha))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {50\vol{10r}}\left(\log({50\vol{10r}}+1) + {2\vol{r}}(1+\log(2|\Phi_\alpha|))\right) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^4+r^2\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d$, this can therefore be satisfied for some $d\in O(r^4)$.

@@

  Therefore, by \cref{weighted_lovasz}, there exists a permutation $\pi$ of $Y$ that produces a bipartition graph $H$ with $\deg_H(\alpha)\ge {2\vol{r}}$ for each $\alpha\in X$.  On the other hand, $\deg_{H}(y)\le |\VE(G_r(y))|-1 < {\vol{r}}$ for each $y\in Y$.  Therefore, for any $A\subseteq X$, $|N_H(A)|\ge {2\vol{r}}|A|/{\vol{r}} = 2|A|$.  Therefore, by \cref{d_hall}, there is a subgraph $M$ of $H$ in which $\deg_M(\alpha)=2$ for each $\alpha\in X$ and $\deg_M(\mu)\le 1$ for each $\mu\in Y$. Each edge $\alpha\mu$ of $M$ corresponds to some $\nu\in X_\alpha\cap\VE(G_r(\mu))$, which we place in $Q_2$.

>>>>

\begin{lem}\label{doubled_colour_set}
  For every integer $r\ge 1$ there exists an integer $d\in O(r^4\log r)$ such that the following is true, for every integer $k\ge 1$.\newline
  Let $G$ be a $k\times k$ pseudogrid with grid-partition $\mathcal{P}:=\{V(P_\mu):\mu\in\VE(G_k)\}$ and let $\varphi$ be a vertex colouring of $G$ with the property that, for each $A\subseteq\varphi(V(G))$, $|\varphi_{\mathcal{P}}^{-1}(A)\cap\interior_r(G_k)|\ge d|A|$.
  Then there exists $S\subseteq V(\interior_r(G))$ such that
  \begin{compactenum}[(i)]
    \item \label{hits_both} for each $\alpha\in\varphi(V(G))$, $|\varphi^{-1}(\alpha)\cap S|= 2$; and
    \item \label{spread_out} for each $v\in S$, $|\tilde{B}_r(v)\cap S|\le 2$ for each $v\in S$.
  \end{compactenum}
\end{lem}

@@

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\ge 2\vol{r}$ for each $\alpha\in X$.
  For each $\alpha\in X$, consider the set $\phi^{-1}_{\mathcal{P}}(\alpha)\subseteq \VE(G_k)$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\phi^{-1}_{\mathcal{P}}(\alpha)| \ge d$.  Furthermore all but at most $\vol{r}$ elements of $\phi^{-1}_{\mathcal{P}}(\alpha)$ are contained in $\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))$ since otherwise the first round would have been an $\alpha$-success.  Let $\Phi_\alpha:=\phi^{-1}_{\mathcal{P}}(\alpha)\cap \left(\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))\right)$, so $|\Phi_\alpha|\ge d-\vol{2r+2}$.  In words, $\Phi_\alpha$ consists of all objects of colour $\alpha$ that could not be chosen in the first round because they are too close to an element of $Y$.

  % \todo{Explain more or redefine $\Phi_\alpha$ as $\Phi_\alpha\cap (\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu)))$)}
  % For each $\mu\in Y$, $|\VE(G_{r+1}(\mu))|\in O(r^2)$. Therefore, there are $\Omega(|\Phi_\alpha|/r^2)$ objects $\mu\in Y$ such that $\VE(G_{r+1}(\mu))\cap \Phi_\alpha\neq\emptyset$.

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\ceil{|\Phi_\alpha|/\vol{10r}},1/25)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 2\vol{r})$.  (We make no attempt to optimize constants here.)  Suppose that, for some $\mu\in Y$, $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\xi\in Y:\nu\in\VE(G_{2r}(\xi))\}
  \]
  has size at most $25$. Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/25$. This is already enough to establish that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$, but in order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.\footnote{There are many ways to obtain a concentration result here, we use a simple method with the goal of being self-contained.  For a tighter result, use the method of bounded differences \cite{mcdiarmid:on}.}  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_r(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J-\VE(G_{10r}(\mu))$.\footnote{The constant $10$ is overkill here and is only used for simplicity; $8r+O(1)$ would be sufficient.}  This process continues until $\VE(G_r(\mu))\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{10r}(\mu))|\le \vol{10r}$, each step in this process eliminates at most $\vol{10r}$ elements from $J$ so the number, $t$, of iterations is at least $\ceil{|\Phi_\alpha|/\vol{10r}}$.

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the  permutation $\pi$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.\footnote{A random variable $X$ dominates a random variable $Y$ if $\Pr(X\ge x)\ge \Pr(Y\ge x)$ for all $x\in\R$.}  Therefore,
  \begin{align}
    \Pr\left(\deg_H(\alpha)< 2\vol{r}\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 2\vol{r}\right) \notag \\
    & \le \Pr(\binomial(t,1/25)\le 2\vol{r}) \notag \\
    & = \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{1}{25}\right)^x\left(\frac{24}{25}\right)^{t-x} \notag \\
    & < \sum_{x=0}^{2\vol{r}} \binom{t}{x}\left(\frac{24}{25}\right)^{t} \notag \\
    & < (2t)^{2\vol{r}}\cdot\left(\frac{24}{25}\right)^{t} \notag \\
    % & \le \exp(6(2r+1)^2\log(2t)+(t-6(2r+1)^2)\log(1-1/25)) \\
    & = \exp\big({2\vol{r}}\ln(2t)-t\ln(25/24)\big) \notag \\
    & \le \exp\left({2\vol{r}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|\ln(25/24)}{\vol{10r}}\right)
      & \text{(since $|\Phi_\alpha|/\vol{10r} \le t \le |\Phi_\alpha|$)} \notag \\
    & \le \exp\left({2\vol{r}}\ln(2|\Phi_\alpha|)-\frac{|\Phi_\alpha|}{25\vol{10r}}\right)
      & \text{(since $\ln(25/24) > 1/25$).}
       \label{probability}
    % & \le \exp(-6(2r+1)^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    % & = \exp(ar^2 - bt)
  \end{align}
  % for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  % $\Pr(\deg_H(\alpha)< 6(2r+1)^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.}   By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

  We are now ready for an application of \cref{weighted_lovasz}.  For each $\alpha\in X$, let $E_\alpha$ denote the event ``$\deg_H(\alpha)\le 2\vol{r}$'' and let $\mathcal{E}:=\{E_\alpha:\alpha\in X\}$. For each $\alpha\in X$, define
  \[
     \Gamma_\alpha := \left\{E_\beta \in \mathcal{E}: \Phi_\beta\cap \left(\bigcup_{\nu\in \Phi_\alpha} \VE(G_{10r}(\nu))\right)\neq\emptyset\right\}
  \]
  and observe that $|\Gamma_\alpha|\le \vol{10r}\cdot|\Phi_\alpha|$.
  The event $E_\alpha$ is mutually independent of $\mathcal{E}\setminus\Gamma_\alpha$ since, for any $E_\beta\in\mathcal{E}\setminus\Gamma_\alpha$ the sets $Y_\alpha:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\alpha\neq\emptyset\}$ and $Y_\beta:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\beta\neq\emptyset\}$ are disjoint and $\deg_H(\alpha)$ and $\deg_H(\beta)$ are each determined by the relative orders of $Y_\alpha$ and $Y_\beta$ in $\pi$, respectively.

  For each $\alpha\in X$, let $w(E_\alpha):=1/\tau$, for some $\tau>0$ to be determined shortly.  Then
  \begin{align}
    w(E_\alpha)\prod_{E_\beta\in\Gamma_\alpha} (1-w(E_\beta))
    & = \tfrac{1}{\tau}\left(1-\tfrac{1}{\tau}\right)^{|\Gamma_\alpha|-1} \notag \\
    & \ge \exp\left(-\ln\tau - |\Gamma_\alpha|/(\tau-1)\right) & \text{(Since $(1-1/\tau)^{\tau-1} > 1/e$)} \notag \\
    & \ge \exp\left(-\ln\tau - \vol{10r}\cdot|\Phi_\alpha|/(\tau-1)\right) \enspace . \label{weight_product}
  \end{align}

  Comparing \cref{weight_product,probability} we see that both quantities decrease exponentially in $|\Phi_\alpha|$ but \cref{weight_product} does so at a rate that can be controlled with $\tau$.
  Taking $\tau = {50\vol{10r}}+1$, and using \cref{weight_product,probability}, we find that
  \[
     \Pr\left(E_\alpha\right) \le w(E_\alpha)\prod_{\beta\in\Gamma_\alpha}(1-w(\Gamma_\alpha))
  \]
  provided that
  \[   |\Phi_\alpha| \ge {50\vol{10r}}\big(\ln({50\vol{10r}}+1) + {2\vol{r}}\ln(2|\Phi_\alpha|)\big) \enspace .
  \]
  The right hand side of this last equation is of the form $O(r^2+r^4\log|\Phi_\alpha|)$.  Since $|\Phi_\alpha|\ge d$, this can therefore be satisfied for some $d\in O(r^4\log r)$.

@@

  Therefore, by \cref{weighted_lovasz}, there exists a permutation $\pi$ of $Y$ that produces a bipartite graph $H$ with $\deg_H(\alpha)\ge {2\vol{r}}$ for each $\alpha\in X$.  On the other hand, $\deg_{H}(y)\le |\VE(G_r(y))|-1 < {\vol{r}}$ for each $y\in Y$.  Therefore, for any $A\subseteq X$, $|N_H(A)|\ge {2\vol{r}}|A|/{\vol{r}} = 2|A|$.  Therefore, by \cref{d_hall}, there is a subgraph $M$ of $H$ in which $\deg_M(\alpha)=2$ for each $\alpha\in X$ and $\deg_M(\mu)\le 1$ for each $\mu\in Y$. Each edge $\alpha\mu$ of $M$ corresponds to some $\nu\in X_\alpha\cap\VE(G_r(\mu))$, which we place in $Q_2$.
